### 2023-09-19 proofs/main.tex
Add Declarative Typing comments

@@

\bigskip
\ottgrammartabular{
  \ottv\ottinterrule
  \ottc\ottinterrule
}

\subsection{Declarative Type Inference}
\ottdefnsDT

@@

\subsection{Algorithmic Type Inference}
\ottdefnsAT

\subsection{Singularity}
\ottdefnsSING

>>>>

In the previous section, we presented the 
type system together with subtyping specification 
and the algorithm. In this section, 
we describe the language under this type system, 
together with the type inference specification
and algorithm.

@@

First, we define the syntax of the language.
The language combines System F with call-by-push-value 
style. 
\begin{definition}[Language Grammar]
  \hfill
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottc\ottinterrule
    }

    \ottgrammartabular{
      \ottv\ottinterrule
    }
  \end{multicols}
\end{definition}
Notice that the language does not have
first-class applications: instead, 
we use applicative let bindings---
constructions that bind a result of
a fully applied function to a (positive) 
variable.
In the call-by-push-value paradigm, 
it corresponds to monadic bind or do-notation. 
Typewise, these let-binders come in two forms:
annotated and unannotated. The annotated let-binders
$[[let x:iP = v(args); c]]$ 
requires the application to infer the annotated
$[[iP]]$, whereas the unannotated 
$[[let x = v(args); c]]$ 
is used when the inferred type is unique.

A computation of a polymorphic type is constructed 
using $[[Λα⁺ . c]]$, however, the elimination of $[[∀]]$
is implicit. Conversely, the existential types
are constructed implicitly and eliminated 
using the standard unpack mechanism:
$[[let∃ (nas, x) = v; c]]$.

Another dual pair of constructions are 
$[[return v]]$ and $[[{c}]]$. 
The former allows us to embed a value in a pure 
computations. The latter, on the contrary,
encapsulates a thunk of computation in a value.

Finally, the language has a number standard constructions:
lambda-abstractions $[[λx:iP.c]]$,
standard let-bindings $[[let x = v; c]]$,
and type annotations that can be added to any value or computation:
$[[(v:iP)]]$ and $[[(c:iN)]]$.

\subsection{Declarative Type Inference}

Next, we define the specification of the type 
inference for our language.
The judgement $[[Γ ; Φ ⊢ v : iP]]$
means that under the type context $[[Γ]]$ and
the variable context $[[Φ]]$, for the value $[[v]]$,
type $[[iP]]$ is inferrable. The negative form of the judgement is symmetric.
The application inference judgement $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ means that if
a head of type $[[iN]]$ is applied to list of values $[[args]]$,
then the resulting computation can be typed as $[[iM]]$.

\begin{definition}[Variable Context]
  The variable typing context $[[Φ]]$
  is represented by a set of entries of the form
  $[[x : iP]]$. 
\end{definition}

\begin{definition}[Declarative Type Inference]
  \hfill
  \begin{multicols}{2}
  \ottdefnDTNInf{}
  \ottdefnDTPInf{}
  \ottdefnDTSpinInf{}
  \end{multicols}
\end{definition}

Let us discuss selected rules of the declarative system:
\begin{itemize}
  \item \ruleref{\ottdruleDTVarLabel}
    says that the type of a variable is inferred from the context.
  \item \ruleref{\ottdruleDTThunkLabel} says
    that the type of a thunk is inferred by shifting up the type of the 
    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
    infers the type of a return by shifting down the type of the
    contained value.
  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
    They allow the inferred type to be refined by annotating it with a supertype.
  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
    mean that th declarative system allows to infer any type from the equivalence class.
  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
    and its first premise inferring the existential type of the value being unpacked.
    It is important however that the inferred existential type is normalized. 
    This is because there might be multiple equivalent existential types 
    with different order or even number of quantified variables, 
    and to bind them, the algorithm needs to fix the canonical one.
  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
    with annotated let-bindings. The first premise infers the type of the head of the application,
    which must be a thunked computation. Then if after applying it 
    to tha arguments, the resulting type can be equated to the annotated one,
    we infer the body of the let-binding in the context extended with the bound variable.
  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
    but it is used when the type of the application is unique, and thus, the annotation
    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
\end{itemize}

Let us discuss the rules of the application inference:
\begin{itemize}
  \item \ruleref{\ottdruleDTEmptyAppLabel} 
    is the base case. If the list of arguments is empty, 
    the inferred type is the type of the head. 
    However, we relax this specification by allowing it to 
    infer any other equivalent type. 
    The relaxation of this rule is enough to guarantee 
    this property for the whole judgement:
    if $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ then 
    $[[Γ ; Φ ⊢ iN ● args ⇒> iM']]$ for any equivalent 
    $[[iM']]$.
  \item \ruleref{\ottdruleDTArrowAppLabel}
    is where the application type is inferred: 
    if the head has an arrow type $[[iQ → iN]]$,
     we are allowed to apply it as soon as 
    as soon as the first argument has a type, which is a subtype of $[[iQ]]$.
  \item \ruleref{\ottdruleDTForallAppLabel}
    is the rule ensuring the implicit elimination of the universal quantifiers. 
    If we are applying a polymorphic computation, 
    we can instantiate its quantified variables with any types,
    which is expressed by the substitution $[[Γ ⊢ σ : {pas}]]$.
\end{itemize}

@@

\subsection{Algorithmic Type Inference}
\begin{algorithm}
  \hfill
  \begin{multicols}{2}
  \ottdefnATNInf{}

  \columnbreak
  \ottdefnATPInf{}
  \ottdefnATSpinInf{}
  \end{multicols}
\end{algorithm}

\subsection{Constraint Singularity}

\begin{algorithm}
  \hfill
  \begin{multicols}{2}
    \ottdefnSINGscEN{}

    \columnbreak
    \ottdefnSINGscEP{}
  \end{multicols}
\end{algorithm}

### 2023-09-19 proofs/tex/equiv-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Add Declarative Typing comments

\begin{corollary}[Normalization preserves subtyping] 
  \label{corollary:nf-pres-subt}
  Assuming all the types are well-formed in context $[[Γ]]$,
  \begin{itemize}
    \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
    \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
  \end{itemize}
\end{corollary}
\begin{proof}
  \hfill
  \begin{itemize}
    \item [$+$]  
    \begin{itemize}
      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
        By \cref{corollary:nf-sound-wrt-subt-equiv},
        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
        in particular, by inversion, 
        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
        Then by the transitivity of subtyping 
        (\cref{corollary:subtyping-transitivity}), 
        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
        and inversion, 
        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
    \end{itemize}
    \item [$-$] The negative case is proved symmetrically.
  \end{itemize}
\end{proof}

@@

\begin{lemma} \label{lemma:mutual-subst-subtyping}
  Asssuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
  are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,
  \begin{itemize}
  \item[$+$] if $[[Γ ⊢ [σ1] iP ≥ iQ]]$ and $[[Γ ⊢ [σ2] iQ ≥ iP]]$\\
    then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iQ]]$;
  \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
    then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iM]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  \hfill
  \begin{itemize}
  \item[$+$]
    Applying $[[σ2]]$ to both sides of
    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{lemma:subst-pres-subt}),
    we have: $[[Γ ⊢ [σ2 ○ σ1] iP ≥ [σ2]iQ]]$.
    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ by transitivity 
    (\cref{lemma:subtyping-transitivity}),
    we have $[[Γ ⊢ [σ2 ○ σ1] iP ≥ iP]]$.
    Then by \cref{corollary:subst-proper-subt},
    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$.
    By a symmetric argument, we also have:
    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$.

>>>>

\begin{corollary}[Normalization preserves subtyping] 
  \label{corollary:nf-pres-subt}
  Assuming all the types are well-formed in context $[[Γ]]$,
  \begin{itemize}
    \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
    \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
  \end{itemize}
\end{corollary}
\begin{proof}
  \hfill
  \begin{itemize}
    \item [$+$]  
    \begin{itemize}
      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
        By \cref{corollary:nf-sound-wrt-subt-equiv},
        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
        in particular, by inversion, 
        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
        Then by transitivity of subtyping 
        (\cref{lemma:subtyping-transitivity}), 
        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
        and inversion, 
        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
    \end{itemize}
    \item [$-$] The negative case is proved symmetrically.
  \end{itemize}
\end{proof}

@@

\begin{lemma} \label{lemma:mutual-subst-subtyping}
  Assuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
  are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,
  \begin{itemize}
  \item[$+$] if $[[Γ ⊢ [σ1] iP ≥ iQ]]$ and $[[Γ ⊢ [σ2] iQ ≥ iP]]$\\
    then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iQ]]$;
  \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
    then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iM]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  \hfill
  \begin{itemize}
  \item[$+$]
    Applying $[[σ2]]$ to both sides of
    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{lemma:subst-pres-subt}),
    we have: $[[Γ ⊢ [σ2 ○ σ1] iP ≥ [σ2]iQ]]$.
    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ by transitivity 
    (\cref{lemma:subtyping-transitivity}),
    we have $[[Γ ⊢ [σ2 ○ σ1] iP ≥ iP]]$.
    Then by \cref{corollary:subst-proper-subt},
    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$.
    By a symmetric argument, we also have:
    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$.
